# Continuous Random Variables
In the last chapter, we learned about discrete random variables, expectations, variance, and some common discrete distributions. In this chapter we will explore similar concepts for random variables with continuous spaces. We also get to learn (yes, I'm excited) about one of the most useful, prolific, and benignly named distributions: The normal distribution.

## Probability Density and the Continuous Random Variable
As you may have guessed, and as I have alluded to, the way we define a continuous random variable is somewhat different from our approach to discrete variables. Continuous random variables are distinguished in part by their **space** - wherein an outcome may take on any value in a *continuous range* of values. We may, for example, define the set of outcomes for a continuous random variable to be all the values between and including 0 and 1, e.g., $0 \leq X \leq 1$, or all the values on the real number line, e.g., $-\infty < x < \infty$. Continuous random variables are prolific within the behavioral sciences: Age, height, and reaction time, for example, can all be thought of as continuous random variables. 

To fully specify a continuous random variable, we shall again need a set of rules describing its randomness, a.k.a., a probability distribution. This is where things get tricky. As you will see the difficulty arises from our aforementioned definition of continuous spaces. Let's explore this by considering a random variable with the space - $0 \leq X \leq 1$. Suppose we would like to define a probability distribution for this random variable. One (un)reasonable thing to try would be to proceed as we would for a discrete random variable. We could take a probability mass of 1 and spread it around over of the values in the range. That is to say, we could try to define a probability mass function for our continuous random variable. 

Take a moment to glance back at the properties of a probability mass function (if you don't remember them). Specifically, over the $n$ outcomes in the space of a discrete random variable $X$, a pmf must have

$$
\sum_{i = 1}^{n} p(x_i) = 1
$$

This property will be quite difficult to attain in the continuous case - because, as we will see, the values in a continuous range aren't countable. That is, we can't list them out in an orderly fashion (e.g., like we could for the Bernoulli, binomial, and poisson distributions). Turning our attention back to our random variable with space $0 \leq X \leq 1$, let's try to count all the values between 0 and 1. Suppose we've listed (or believe we have listed) all the values in the range, e.g.,

$$
0.123451..., \:0.234672...,\:0.247773...,\:0.866724...,\:0.999525...
$$

We line our list up with the natural numbers and begin counting. 

$$
1) \:\: 0.123451...\\
2) \:\: 0.234672...\\
3) \:\: 0.247773...\\
4) \:\: 0.866724...\\
5) \:\: 0.999525...
$$

But to our surprise (or maybe to our non-surprise) there will *always* be a number we missed during our enumeration. Note the underlined digits in our list below.

$$
1) \:\: 0.\underline{1}23451...\\
2) \:\: 0.2\underline{3}4672...\\
3) \:\: 0.24\underline{7}773...\\
4) \:\: 0.866\underline{7}24...\\
5) \:\: 0.9995\underline{2}5...
$$

We can specify a new number $0.abcde$ such that a, b, c, d, and e are numbers between 0 and 9, representing the first through fifth digits, respectively. To make sure our number is new, we ensure that $a \neq 1$, $b \neq 3$, $c \neq 7$, $d \neq 7$, and $e \neq 2$. Thus the first digit will differ from that of the first listed number, the second digit will differ from that of the second listed number, and so on. Our number is therefore different from each of the numbers in our list. This method of finding a new number would *always work*. No matter how certain we were in the thoroughness of our list, we would always have to go back and account for a newly discovered number.

The logic described above, credited to Georg Cantor a famous set theorist, is indeed intriguing (in fact it's one of my favorite proofs of all time!), but why is it important for our purposes? It tells us that we can't specify probability distributions for continuous random variables in the way that we define them for discrete random variables. You can't sum the probability of each outcome - $p(x_i)$ - over the set of outcomes _because we can't list the outcomes out_. And, therefore, we can't ensure we've attained the important property of $\sum_{i = 1}^{n} p(x_i) = 1$.  

As it turns out, we get around this conundrum by assigning probabilities to *ranges* of our continuous random variable, and we frequently do this with something called a **probability density function** (pdf). A probability density function describes the amount of probability (sometimes referred to here as *probability mass*) _per unit_ $x$ at a given point. As you will see, the probability density will help us determine the probability of an outcome falling within a given range. For example, it could tell us the probability that $X$ falls below the value of $.5$, e.g., $P(X < .5)$. We'll dig deeper into this concept as we have in the past - through an example. 

::: {.example #firstdensity}

Suppose we have a probability density function

$$
f(x) = 
\begin{cases}
1 & 0 \leq x \leq 1 \\
0 & Everywhere \: else
\end{cases}
$$

This function assigns a value of 1 probability density to any value - $0 \leq x \leq 1$. Anywhere else it assigns a value of 0. Let's refer to units of probability mass (or probability) with the abbreviation - $pm$ - and let's assume that the units of $x$ are $u$ (short for units of x). Suppose our value of $x = .25$, then

$$
f(.25) = 1 
$$

The value of 1 from our pdf is a probability density which has the units $pm/u$, e.g., probability mass per units of x. That is to say, our probability density function is describing the amount of probability mass per unit $x$ at exactly the point $x = .25$. Now let's talk about how such a function assigns probability to a range of x. Take a look at the plot of our probability density function below.

```{r, echo = FALSE, fig.cap = "Boring"}
a <- data.frame(x = c(0,1), 
                y = c(1,0),
                x_end = c(1,1.5),
                y_end = c(1,0))

b <- data.frame(x = c(0,1), 
                y = c(1,1))

d <- data.frame(x = c(1), 
                y = c(0))
                

ggplot2::ggplot(a, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_segment(data = a, ggplot2::aes(x = x, y = y, xend = x_end, yend = y_end)) +
  ggplot2::scale_y_continuous(limits = c(0,2)) +
  ggplot2::scale_x_continuous(limits = c(0,1.5)) +
  ggplot2::labs(y = "y (pm/u)", x = "x (u)") +
  ggplot2::geom_point(data = b, ggplot2::aes(x=x,y=y)) +
  ggplot2::geom_point(data = d, ggplot2::aes(x=x,y=y), shape = 21, fill = 'white')


```

Indeed, it is a very boring plot - just a horizontal line where $f(x) = 1$ from 0 to 1. We're going to calculate an area under that line, and that area will be a probability mass! For demonstration sake, let's stick with $x = .25$ and calculate the area under the $f(x)$ line from $0 \leq x \leq .25$. Take a look at the the rectangle with the diagonal lines through it in the plot below. The area to which we are referring is bordered by $f(x)$ at the top, the x-axis at the bottom, and vertical lines at $x = 0$ and $x = .25$, respectively. 

```{r, echo = FALSE, fig.cap = "Area under the density"}

a <- seq(-.25,1, by = .05)
b <- data.frame(x = c(0,.25,0,0,1), 
                y = c(1,0,0,0,0),
                x_end = c(1,.25,.25,0,1.5),
                y_end = c(1,1,0,1,0))

df <- data.frame(x = ifelse(a < 0, abs(a), 0), 
                 y = ifelse(a < 0, 0, a), 
                 x_end = ifelse(a + .25 > 1, (1-a), .25),
                 y_end = ifelse(a + .25 > 1, 1, a + .25))

b <- data.frame(rbind(b,df))


c <- data.frame(x = c(0,1), 
                y = c(1,1))


d <- data.frame(x = c(1), 
                y = c(0))
                

ggplot2::ggplot(b, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_segment(data = b, ggplot2::aes(x = x, y = y, xend = x_end, yend = y_end)) +
  ggplot2::scale_y_continuous(limits = c(0,2)) +
  ggplot2::scale_x_continuous(limits = c(0,1.5)) +
  ggplot2::labs(y = "y (pm/u)", x = "x (u)") +
  ggplot2::geom_point(data = c, ggplot2::aes(x=x,y=y)) +
  ggplot2::geom_point(data = d, ggplot2::aes(x=x,y=y), shape = 21, fill = 'white')


```

Recall that the equation for calculating the area of a rectangle is the length of the base (call it $B$) times the height (call it $H$) - $BH = area$. Here the length of the base is $.25$, and the height of our rectangle is $f(.25)=1$. Also note the units: The length has units - $u$ (units of x) - and the height has units of probability density - $pm/u$ - probability mass per unit $x$. Therefore

$$
area \: = \: BH \: = \frac{.25\:u}{1} \times \frac{1\:pm}{1\:u} = .25 pm  
$$

We get an area of .25, but the story of the calculation is in the units of this value. Note how the units of $x$ cancel out, and the units of our calculated value are $pm$ (probability mass). That is to say, our probability density function has assigned a 25% probability to the range $0 \leq x \leq .25$. Put another way, we have $P(0\leq X\leq.25) = .25$. Indeed, if we calculate the area under the whole curve (line) where $f(x) = 1$, then we would find that the area is 1. That is, there is a 100% chance that our random variable $X$ falls in the range $0 \leq x \leq 1$. This mimics the property of pmfs we discussed at the start of this section.

:::

There we have it, then. The probability density function assigns a probability to a range via the area under its curve. In practice, calculation of this area requires more advanced mathematics than we have demonstrated above. For example, consider the function

$$
f(x) =
\begin{cases}
e^{-x} & 0 \leq x \leq \infty \\
0 & x<0
\end{cases}
$$
This function is an example of an *exponential distribution*. When we look at its plot (below) it becomes apparent that calculating the area under it's curve is a bit more complicated. 

```{r echo = FALSE}
x <- seq(0,5,.05)
y <- exp(-x)
z <- data.frame(x,y)

ggplot2::ggplot(data = z, ggplot2::aes(x=x,y=y))+
  ggplot2::geom_point() +
  ggplot2::geom_line()
```

We can't just calculate this area as we did with our rectangular region. More advanced techniques (that you would learn in Calculus 2) do show that the area under the curve of this function is equal to 1. That is, $P(0 < X < \infty) = 1$. The means by which we calculate the area under a given probability density function (and a function more generally) is called **integration**. The concept of integration is well beyond the scope of this book, so we shall only discuss it briefly and at high level. If you are excited enough to about the concept of integration, you are certainly capable of learning it in more detail. 

Given some function, $f(x)$, the operation of integration is denoted

$$
area \:=\int_{a}^{b} f(x) \,dx
$$

The values a and b indicate the starting point and the ending point for the integration. They tell us to calculate the area under the curve $f(x)$ from a to b. In our first example above, we calculated the integral of $f(x)$ from $x = 0$ to $x = .25$. In terms of our probability set operator, and given some pdf, we observe that

$$
P(a < X < b) = \int_{a}^{b} f(x) \, dx
$$
That is to say, in the context of a pdf (and a continuous random variable), the probability that the outcome of some random experiment falls in a given range from a to b is the integral (area) of the pdf from a to b. Indeed for our purposes, it is important for you to know this at a high level. You shall not - except where your skills will allow - be asked to do an integral in this book. However, it's helpful to know what an integral is and what a probability density is. It will make our foray into continuous distributions much more interesting. 

Next we turn our attention to the properties of a pdf, which are analogous to the discrete case. Specifically

1. For all values of $x$ in the space of $X$, $f(x) \geq 0$. The values for which $f(x) > 0$ are referred to as the **support** of the continuous random variable. This is analogous to the discrete case and pmf. However, note the subtle difference: $f(x)$ can be greater than 1. 

2. $\int_{-\infty}^{\infty} f(x) \, dx = 1$. Put in words, the area under the pdf is equal to 1. Here, we use negative infinity and positive infinity for the integral to make things easier notation wise. Any value outside the space of the random variable $X$ is assigned a probability density of zero. That is to say, the ranges of "infinity" outside the space of $X$ have probability zero, and therefore, writing negative and positive infinity doesn't change the outcome of the integral/area calculation. 

Now, let's return to our simple pdf from the beginning of this section

$$
f(x) = 
\begin{cases}
1 & 0 \leq x \leq 1 \\
0 & Everywhere \: else
\end{cases}
$$

Note that this function has all the properties of a pdf - $f(x) \geq 1$ and $\int_{-\infty}^{\infty} f(x) \, dx = 1$. The area under $f(x)$ from 0 to 1 is equal to 1, and everywhere else the area is equal to zero. That is to say, over the entire space of $X$ the area under the curve is equal to 1. This particular function is an example of a *continuous uniform distribution*. These functions assign a constant probability density over the support of $X$ and are analogous to the discrete uniform distribution. Here's another example that assigns $\frac{1}{2}$ density from 0 to 2.

$$
f(x) = 
\begin{cases}
1/2 & 0 \leq x \leq 2 \\
0 & Everywhere \: else
\end{cases}
$$

### Practice Problems

For the next few problems describe if the function is *not* a pdf and explain why. It will be helpful to draw a plot of each function and recall (or look-up) calculations for geometric shapes.

1. $f(x) = x$ from $-1 < x < 1$ and zero everywhere else.

2. $f(x) = 2$ from $0 < x < 1/2$ and zero everywhere else.

3. $f(x) = x$ from $0 < x < \sqrt{2}$ and zero everywhere else (Hint: draw the plot).

Given the pdf below. For problems 4-7 calculate the probability that an outcome of a random experiment falls in the specified range.
$$
f(x) = 
\begin{cases}
1/2 & 0 \leq x \leq 2 \\
0 & Everywhere \: else
\end{cases}
$$

4. What is $P(0 < X < .25)$?

5. What is $P(1 < X < 2)$?

6. What is $P(.25 < X < 1)$?

7. What is $P(X > 2)$

8. This problem may be tough. Don't get discouraged. Given the pdf

$$
f(x) = 
\begin{cases}
x & 0 \leq x \leq \sqrt{2} \\
0 & Everywhere \: else
\end{cases}
$$

Describe a function - $F(x)$ - that takes a value of $x$ and returns $P(0 < X < x)$. Think about the area of a triangle.

### Practice Problem Answers
1. $f(-1) = -1$, so this function can't be a pdf because $-1 < 0$.
2. This function would work as a probability density.
3. This function would work as a probability density.
4. 1/8
5. 1/2
6. 3/8
7. 0
8. $\frac{1}{2}xf(x) = \frac{1}{2}xx = \frac{1}{2}x^2$. If you got that part, well done! If you didn't, no worries it was tricky. Now there's a subtle point that must be made which I would not have fully expected you to notice (because of my explanations; not because of any shortcomings you may have). The density is zero after $\sqrt(2)$ and before zero. After $\sqrt(2)$ there is zero area under the density curve and therefore zero additional probability added. In other words, 

$$
F(x) = 
\begin{cases}
0 & x < 0 \\
\frac{1}{2}x^2 & 0 \leq x \leq \sqrt{2} \\
1 & \sqrt{2} < x
\end{cases}
$$

## Cumulative Distribution Functions, Independence, and Expectations

### Cumulative Distribution Functions

As practitioners, we usually won't work directly with the probability density of a continuous random variable. Instead, we will work with something called a **cumulative distribution function** (cdf). Given a random variable with a probability density function, $f(x)$, a cdf determines the area under the density curve left of (less than) a given value of $x$. That is to say, it tells us $P(X < x)$. We often refer to these functions with a capital $F$ to distinguish them from their corresponding pdf $f$. In other words,

$$
F(x) = P(X < x)
(\#eq:cdfs)
$$
Note that you may also see $F(x) = P(X \leq x)$ in other literature. This is an equivalent way of writing the equation above.

::: {.example #cumulative}
In order to get a better understanding of cdfs, it helps to derive one. We don't want to get involved with complicated integrals, so let's return to our first ever probability density (it was so long ago!). 

$$
f(x) = 
\begin{cases}
1 & 0 \leq x \leq 1 \\
0 & Everywhere \: else
\end{cases}
$$

We would like to determine the cdf for our random variable $X$, and for that purpose, it helps to look at a graph and calculate a few values. Take a look at the very boring graph of our pdf again.

```{r, echo = FALSE, fig.cap = "Very Boring"}
a <- data.frame(x = c(0,1), 
                y = c(1,0),
                x_end = c(1,1.5),
                y_end = c(1,0))

b <- data.frame(x = c(0,1), 
                y = c(1,1))

d <- data.frame(x = c(1), 
                y = c(0))
                

ggplot2::ggplot(a, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_segment(data = a, ggplot2::aes(x = x, y = y, xend = x_end, yend = y_end)) +
  ggplot2::scale_y_continuous(limits = c(0,2)) +
  ggplot2::scale_x_continuous(limits = c(0,1.5)) +
  ggplot2::labs(y = "y (pm/u)", x = "x (u)") +
  ggplot2::geom_point(data = b, ggplot2::aes(x=x,y=y)) +
  ggplot2::geom_point(data = d, ggplot2::aes(x=x,y=y), shape = 21, fill = 'white') +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::theme_bw()

```

Suppose we want to calculate $P(X < x)$ for several specified values of $x$, say, .25, .5, .75, and 1. For the purposes of clarity, let's call each of these values $x_1$, $x_2$, $x_3$ and $x_4$, respectively. In the figure below, I've drawn a vertical line corresponding to each $x_i$. To calculate $P(X < x_i)$, we will need to determine the area under the pdf from $-\infty$ up to the line corresponding to the value of $x_i$. When we are done with our calculations, we will have a better idea of the cdf of $x$.

```{r, echo = FALSE, fig.cap = "Very Very Boring"}
a <- data.frame(x = c(0,1), 
                y = c(1,0),
                x_end = c(1,1.5),
                y_end = c(1,0))

b <- data.frame(x = c(0,1), 
                y = c(1,1))

d <- data.frame(x = c(1), 
                y = c(0))

e <- data.frame (x = c(.25,.5,.75,1), 
                y = c(0,0,0,0),
                x_end = c(.25,.5,.75,1),
                y_end = c(1,1,1,1))

f <- data.frame(x = c(.25,.5,.75,1),
                y = c(1.125,1.125,1.125,1.125),
                text = c('x1', 'x2' ,'x3', 'x4'))
                

ggplot2::ggplot(a, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_segment(data = a, ggplot2::aes(x = x, y = y, xend = x_end, yend = y_end)) +
  ggplot2::geom_segment(data = e, ggplot2::aes(x = x, y = y, xend = x_end, yend = y_end)) +
  ggplot2::geom_text(data = f, ggplot2::aes(x = x, y = y, label = text)) +
  ggplot2::scale_y_continuous(limits = c(0,2)) +
  ggplot2::scale_x_continuous(limits = c(0,1.5)) +
  ggplot2::labs(y = "y (pm/u)", x = "x (u)") +
  ggplot2::geom_point(data = b, ggplot2::aes(x=x,y=y)) +
  ggplot2::geom_point(data = d, ggplot2::aes(x=x,y=y), shape = 21, fill = 'white') +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::theme_bw()

```

Thankfully, in this case, we won't need any fancy integration techniques to calculate a couple of values. For the region $0 \leq X \leq 1$, we simply need to multiply the length of the base of each of the rectangles (in this case $x_i$) in the figure above by its height (in this case, $1$). That is to say, we can proceed as in example \@ref(exm:firstdensity). But before we do that, note that we are calculating the area under the pdf for *all* values less than $x$. For example, in the case of $x_1 = .25$, we will need to calculate the area under the curve from $-\infty$ all the way up to the value of $x_i$ which is, unfortunately, not a rectangle (oh, the horror!). Thankfully, our pdf says that any value less than 0 has a probability density of 0, e.g., $f(x) = 0$, so we can safely assume that the area under the curve for any value less than zero is zero.   

We shall therefore proceed as we did in example \@ref(exm:firstdensity). The calculations are written out in the table below.

```{r, echo = FALSE}
ranges <- c('P(X < .25)','P(X < .50)','P(X < .75)','P(X < 1)')
x <- c(.25,.5,.75,1)
y <- c(1,1,1,1)
q <- x * y

sample_space <- data.frame(ranges, x,y,q)
names(sample_space) <- c('Range','Base: x', 'Height: f(x)','P(X < x): Base x Height')

x <- flextable::flextable(sample_space)
x <- flextable::align(x, align = 'center', part = 'body')
x <- flextable::align(x, align = 'center', part = 'header')
x <- flextable::bg(x, bg = 'gray', part = 'header')
x <- flextable::hline(x)
x <- flextable::autofit(x)
x
```

Now, let's plot each of our values of $x$ and it's corresponding $P(X < x)$. Ultimately this will help us get a sense for the cdf of $f(x)$. In this case, we have four $(x,y)$ coordinates - $(.25, P(X < .25))$, $(.50, P(X < .50))$,$(.75, P(x < .75))$, and $(1, P(X < 1))$. If we plot each and connect each with a line, we get

```{r, echo = FALSE, fig.cap = "The (almost) cdf of f(x)"}
a <- data.frame(x = c(0,.25,.5,.75,1), 
                y = c(0,.25,.5,.75,1))

ggplot2::ggplot(a, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point(shape = 21, color = 'black', fill = 'black') +
  ggplot2::geom_line() +
  ggplot2::xlim(c(0,2)) +
  ggplot2::ylim(c(0,2)) +
  ggplot2::theme_bw() +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::labs(y = "P(X < x)", x = "x")

```

Looking at the plot above, and considering our earlier computations, we might begin to think that $P(X < x) = x$. That is to say, the probability that $P(X < x_i)$ is $x_i$. This assertion would be correct. For example, given the values $1/2$, $1/3$, and $57/58$, their probabilities would be

$$
P(X < 1/2) = 1/2 \\
P(X < 1/3) = 1/3 \\
P(X < 57/58) = 57/58
$$
But, unfortunately for us, we've only focused on values in the region $0 \leq X \leq 1$. What shall we say about $P(X < x)$ for values outside this region? Let's first consider a scenario where the value of $x$ is less than zero. Take a look back at our pdf. For any value of $x$ less than 0, the probability density at that point is zero. This means that there is no area underneath the pdf at all points below zero. Put another way, this means that if we were to conduct a random experiment with the $X$ in our example, the outcome would never be less than zero. Thus, $P(X < x)$ is 0 when the value of $x$ is less than zero.

Now suppose $x > 1$, what would we say about $P(X < x)$? Looking back again at our example pdf, notice how all the density is centralized on the values from 0 to 1. Thus, if we were to conduct a random experiment with this random variable, we would be guaranteed a value in that range. All the probability mass (area) is located in that region. Therefore, given some value larger than 1, our random experiment is headed for success. As such, in the case that $x > 1$, we will say that $P(X < x)$ = 1.

We therefore may write our cumulative distribution function like this:

$$
P(X < x) = F(x) = 
\begin{cases}
0 & x < 0 \\
x & 0 \leq x \leq 1 \\
1 & x > 1
\end{cases}
$$

For any value below 0, $f(x)$ has zero density (the height of any rectangle in that region is zero), so $P(X < 0) = 0$. On the other hand, if our value of $x$ is greater than 1, then $P(X < x) = 1$. A random experiment is guaranteed to result in a value less than or equal to one since all the probability density is centralized in $0 \leq X \leq 1$. For all other values, $P(X < x) = x$ 

Our new-found cdf makes computations easier. For example, given the values 0.11, 0.512, 0.98,, -100, 10 then

$$
P(X < 0.11) = F(.11) = 0.11 \\
P(X < 0.512) = F(0.512) = 0.512 \\
P(X < 0.98) = F(0.98) = 0.98 \\
P(X < -100) = 0 \\
P(X < 10) = 1
$$
:::

In practice, we will sometimes want to know the probability of $X$ falling in a range dissimilar from the cdf. We may want to know, for example, $P(a < X < b)$ or $P(X > a)$. In our next examples, we frame the intuition of a some common scenarios. Now may be the time to review sets for continuous sample spaces if you've forgotten.

The examples we work through shall use a continuous random variable - $X$ - with an exponential pdf (a.k.a., an exponential distribution), e.g., 

$$
f(x) = 
\begin{cases}
e^{-x} & x \geq 0 \\
0 & x < 0 
\end{cases}
$$

The exponential pdf has the cdf 

$$
P(X < x) = F(x) = 
\begin{cases}
1-e^{-x} & x \geq 0 \\
0 & x < 0 
\end{cases}
$$

You can see the pdf and the cdf plotted side by side (respectively) in the figure below. Note how the area under the curve for the pdf declines as $x$ gets larger. For example, there's more area under the pdf from 0 to 2.5 than there is from 2.5 to 5. Intuitively speaking, this explains the shape of the cumulative distribution function on the right. If we are calculating $P(X < x)$, incremental increases in $x$ near zero are going to result in relatively large increases in the area under the curve to the left of $x$. Thus the cdf - which represents $P(X < x)$ - increases rapidly for values of $x$ that are closer to zero. Then as $x$ gets larger, incremental increases don't add much area, so $P(X < x)$ grows much slower until it levels out. At very large values of $x$ it will be very close to 1.

```{r, echo = FALSE, fig.cap = "The pdf and cdf for our example"}
x <- seq(0,10, by = .1)
y <- rep(c("f(x)","P(X < x) = F(x)"), each = length(x))
pdf <- exp(-x)
cdf <- 1-exp(-x)
outs <- c(pdf,cdf)
plot_data <- data.frame(outs, y, x)

ggplot2::ggplot(plot_data, ggplot2::aes(x = x, y = outs)) +
  ggplot2::geom_line() +
  ggplot2::theme_bw() +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::facet_wrap(~ y) +
  ggplot2::labs(y = 'Density - f(x)')

```

::: {.example #exp}
Suppose we are working with our previously described exponentially distributed random variable, and we would like to know the probability that the outcome of a random experiment falls between 2 and 3. That is, we would like to know $P(2 < X < 3)$. Using our cdf, we could easily calculate $P(X < 3)$ but that would include all the probability mass in the region below 2. That's not what we want, but we can still work with our cdf to get what we need. In the solution that follows, we shall approach the problem in two ways - one using set notation and another that leans on intuition and graphs.

First, take a close look at the event $X < 3$. It can actually be written as the union of two mutually exclusive events, e.g.,

$$
\{X < 3\} = \{X < 2\} \cup \{2 \leq X < 3\}
$$

This means that we can use the law of addition \@ref(eq:addlaw) to get what we need! To see how, notice that

$$
P(\{X < 2\} \cup \{2 \leq X < 3\}) = P(X < 3) \\
P(X < 2) + P(2 \leq X < 3) = P(X < 3)
$$
Where the second line follows from the law of addition. Further note I dropped the curly braces in the second line for the sake of simplicity. We can now subtract $P(X < 2)$ from both sides and voila

$$
\begin{align}
P(2 \leq X < 3) &= P(X < 3) - P(X < 2) \\
P(2 \leq X < 3) &= F(3) - F(2) \\
P(2 \leq X < 3) &= 1-e^{-3} - (1 - e^{-2}) \\
P(2 \leq X < 3) &= 1-e^{-3} - 1 + e^{-2} \\
P(2 \leq X < 3) &= e^{-2}-e^{-3} = 0.08555
\end{align}
$$

Indeed, in general, given some $a$ and $b$, $P(a < X < b)$ can be calculated by

$$
P(a < X < b) = P(X < b) - P(X < a) = F(b) - F(a)
(\#eq:fundamental)
$$
Note that a and b were 2 and 3 in our example, respectively. An astute student (and I believe you are an astute student) may notice that in our example we calculated $P(2 \leq X < 3)$, and we have just provided an equation for $P(2 < X < 3)$. However, recall our brief discussion at the start of this section - $P(a \leq X < b) = P(a < X < b) = P(a < X \leq b) = P(a \leq X \leq b)$. .

Now that we've worked the problem out with set theory, we shall turn our attention to an intuitive approach. The gray colored region in the plot below is the area under the exponential pdf left of the value 3 (e.g., $X < 3$). As we have learned, it represents $P(X < 3)$. The area between the two vertical lines at the values 2 and 3 under the curve represents $P(2 < X < 3)$, and the area to the left of 2 represents $P(X < 2)$. Looking at the two regions, one can see that $P(X < 2)$ and $P(2 < X < 3)$ combine (sum) to $P(X < 3)$. We shall need to subtract out the area left of 2 to get our desired value.

```{r, echo = FALSE, fig.cap = "P(X < 3)"}
x <- seq(0,10, by = .1)
linx <- seq(0,3, by = .01)
pdf <- exp(-x)
Area <- ifelse(linx < 2, 'P(X < 2)', 'P(X < 3)')
plot_data <- data.frame(pdf, x)
verts <- data.frame(x = c(2,3), y = c(0,0),
                        xend = c(2,3), yend = exp(-c(2,3)))


lins <- data.frame(x_start = linx, y_start = 0, x_end = linx, y_end = exp(-linx),
                   Area)

ggplot2::ggplot(plot_data, ggplot2::aes(x = x, y = pdf)) +
  ggplot2::theme_bw() +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::geom_segment(data = lins, ggplot2::aes(x = x_start, 
                                                  y = y_start, 
                                                  xend = x_end, 
                                                  yend = y_end),
                        color = 'gray') +
  ggplot2::geom_segment(data = verts, ggplot2::aes(x = x, y = y,
                        xend = xend, yend = yend)) +
  ggplot2::geom_line() +
  ggplot2::labs(y = 'Density - f(x)') +
  ggplot2::annotate('segment',x = 3, y = .25, xend = 2.5, yend = 0.05,
                    arrow = ggplot2::arrow(length = ggplot2::unit(0.3, "cm"))) +
  ggplot2::annotate('text', x = 3, y = .3, label = 'P(2 < X < 3)') +
  ggplot2::annotate('segment',x = 2, y = .5, xend = 1, yend = 0.25,
                    arrow = ggplot2::arrow(length = ggplot2::unit(0.3, "cm"))) +
  ggplot2::annotate('text', x = 2, y = .55, label = 'P(X < 2)') +
  ggplot2::scale_x_continuous(breaks = 0:10)

```
That is to say, we will need to subtract $F(2)$ from $F(3)$ to get the area in the region corresponding to $P(2 < X < 3)$
:::

The equation we learned in the last example will be important for our work with cdfs. Let's dig into a couple more scenarios.

::: {.example}
Let's continue using our aforementioned pdf/cdf. Suppose we would like to know the probability that $P(X > 3)$. One may (yet again!) think we are in a conundrum because of the form of our cdf. Nevertheless, we can again use set notation to make our approach clear.

Note that the support of $X$ is $X \geq 0$, and therefore, $P(X \geq 0) = 1$. That is to say, a random experiment would have a 100% chance of yielding a value greater than or equal to 0 because our pdf has all it's support on zero and above. Also, observe that

$$
\{X > 0\} = \{0 \leq X \leq 3\} \cup \{X > 3\} 
$$
In English: $\{X > 0\}$ is the union of two mutually exclusive sets - $0 \leq X \leq 3$ and $X < 3$. We can once again use the law of addition to get our desired probability.

$$
\begin{align}
\{X > 0\} &= \{0 \leq X \leq 3\} \cup \{X > 3\} \\
P(\{X > 0\}) &= P(\{0 \leq X \leq 3\} \cup \{X > 3\}) \\
1 &= P(0 \leq X \leq 3) + P(X > 3) \\
1 - P(0 \leq X \leq 3) &= P(X > 3)
\end{align}
$$
This means that all we need is $P(0 \leq X \leq 3)$ and we will be able to calculate our desired probability. Looking at the pdf of $X$, one can see that there is no area below the curve for $X$ less than zero. In other words, the area under the pdf from 0 to 3 would be equivalent to the area under the curve from three all the way down to $-\infty$. Put another way, $P(0 \leq X \leq 3)$ is equal to $P(X < 3)$, and we know the latter because of our cdf. Specifically 

$$
\begin{align}
P(X > 3) &= 1 - P(0 \leq X \leq 3) \\
P(X > 3) &= 1 - F(3) \\
P(X > 3) &= 1 - (1 - e^{-3}) \\
P(X > 3) &= 0.04978  
\end{align}
$$
Indeed, note the similarity of this approach to example \@ref(exm:poissoninf) with the countably infinite space of the Poisson random variable. Our approach to this example can be rephrased generally to

$$
P(X > x) = 1 - F(x)
(\#eq:infinitycont)
$$
:::

Although we have derived a couple of equations together, you are hopefully beginning to get a sense for the *process* of using the cdf. We can, in general, find the area under the curve of a pdf by adding or subtracting appropriate values of the cdf. You can determine what to add and subtract by using set notation or by visualizing the regions under consideration. Let's hammer this process home with one more example.

::: {.example}
Suppose we are interested in the probability that our random experiment falls below 1 or above three. That is to say, we are interested in the event - $\{X < 1\}$ OR $\{X > 3\}$. The two probabilities are depicted in the figure below by the gray area under our exponential pdf.

```{r, echo = FALSE, fig.cap = "P(X < 2) or P(X > 3)"}
x <- seq(0,10, by = .1)
linx <- seq(0,10, by = .01)
pdf <- exp(-x)
Area <- ifelse(linx < 2, 'P(X < 2)', 'P(X < 3)')
plot_data <- data.frame(pdf, x)
verts <- data.frame(x = c(2,3), y = c(0,0),
                        xend = c(2,3), yend = exp(-c(2,3)))


lins <- data.frame(x_start = linx, y_start = 0, x_end = linx, y_end = ifelse(linx < 2 | linx > 3, exp(-linx),0),
                   Area)

ggplot2::ggplot(plot_data, ggplot2::aes(x = x, y = pdf)) +
  ggplot2::theme_bw() +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::geom_segment(data = lins, ggplot2::aes(x = x_start, 
                                                  y = y_start, 
                                                  xend = x_end, 
                                                  yend = y_end),
                        color = 'gray') +
  ggplot2::geom_segment(data = verts, ggplot2::aes(x = x, y = y,
                        xend = xend, yend = yend)) +
  ggplot2::geom_line() +
  ggplot2::labs(y = 'Density - f(x)') +
  ggplot2::annotate('segment',x = 3, y = .25, xend = 3.5, yend = 0.01,
                    arrow = ggplot2::arrow(length = ggplot2::unit(0.3, "cm"))) +
  ggplot2::annotate('text', x = 3, y = .3, label = 'P(X > 3)') +
  ggplot2::annotate('segment',x = 2, y = .5, xend = 1, yend = 0.25,
                    arrow = ggplot2::arrow(length = ggplot2::unit(0.3, "cm"))) +
  ggplot2::annotate('text', x = 2, y = .55, label = 'P(X < 2)') +
  ggplot2::scale_x_continuous(breaks = 0:10)

```
Our intuitive approach suggests that we can simply add the two areas up to get our desired probability. That is,

$$
\begin{align}
P(X < 2) + P(X > 3) &= F(2) + (1 - F(3)) \\ 
&= 1- e^{-2} + (1 - 1 - e(-3)) = 0.81497
\end{align}
$$
Indeed, the law of addition supports our assertion because ${X < 2}$ and ${X > 3}$ are mutually exclusive sets. In other words, we can add their corresponding probabilities. 
:::

### Independence of Continuous Random Variables
The intuition of independence in the context of continuous random variables is quite similar to that of discrete random variables. A thorough look into the dependence/independence of continuous random variables would require a little too much integration for our needs. Instead, we'll frame an intuition then talk about (the ease of) calculating probabilities with independent continuous random variables. Later in this book, we'll discuss dependence in the context of linear correlation.

Take a moment to recall our discussion of independence in the context of discrete random variables. In that exploration, we found that two variables - e.g., $X_1$ and $X_2$ - are dependent if the probability of a particular outcome in the space of $X_1$ changes with an outcome in the space of $X_2$ (or visa versa). For instance, we discussed the sum of two dice and how the value of the first roll ($X_1$) can impact the probability of getting certain sums ($X_2$).

The intuition for dependence in the case of the continuous random variable is similar. That is to say, given two dependent continuous random variables - $X_1$ and $X_2$ - the cdf of $X_1$ will change given an outcome in the space of $X_2$. Indeed, with the powers of integration, it is possible to calculate and visualize how the cdf of $X_1$ changes given some outcome value of $X_2$. That is, we can calculate $P(X_1 < x_1 | X_2 = x_2)$. With that said, I would like to formally request that you hold onto your mathematical shorts. In the next example we are going to do just that (except we'll skip the actual integration part).

::: {.example} 
Now, suppose we are given two dependent random variables - $X_1$ and $X_2$. Further suppose that $X_2$ has all it's support between and including 0 and 1. Skipping the steps to integration (and ignoring the pdf of $X_1$ and $X_2$), let's assume the cdf of $X_1$ given an outcome of $X_2$ looks like this

$$
P(X_1 < x_1 | X_2 = x_2) = F(x_1|x_2) = 
\begin{cases}
0 & x_1 < 0\\
\frac{x_1^2 + 2x_1x_2}{1 + 2x_2} & 0 \leq x_1 \leq 1\\
1 & x_1 > 1
\end{cases} 
$$
Now, this may be a bit overwhelming at first (it's your first continuous conditional probability after all), but we'll unpack it. First look to the left of the equal sign before the curly brace. Recall that the $|$ in the context of conditional probability means "given". $P(X_1 < x_1 | X_2 = x_2)$ and $F(x_1|x_2)$ are just two different ways of referencing the cdf of $X_1$ *given* some value $x_2$. Next, take a look to the right of the curly brace. As we discussed in the last section, these are the rules for the conditional cdf of $X_1$. The rule in the top row says that there is zero chance that $X_1$ falls below 0. The rule in the bottom row says that there is a 100% chance that $X_1$ falls below any value greater than 1. Put another way, a random experiment with $X_1$ is guaranteed to fall at or below 1. Now take a look at the rule in the middle row. Notice how there are two variables referenced - $x_1$ and $x_2$. It is this rule that will tell us how the cdf of $X_1$ *changes* with specified outcomes of $X_2$. We'll choose a value $0 \leq x_2 \leq 1$ and plug it into the middle rule. For example, let's take the values $x_2 = 0$ and $x_2 = 1$. Thus, for $x_2 = 0$

$$
P(X_1 < x_1 | X_2 = 0) = F(x_1|0) = 
\begin{cases}
0 & x_1 < 0\\
\frac{x_1^2 + 2x_1(0)}{1 + 2(0)} = x_1^2 & 0 \leq x_1 \leq 1\\
1 & x_1 > 1
\end{cases}
$$
That is, we set $x_2 = 0$ and the middle rule changed to $x^2$. On the other hand, for $x_2 = 1$

$$
P(X_1 < x_1 | X_2 = 1) = F(x_1|1) = 
\begin{cases}
0 & x_1 < 0\\
\frac{x_1^2 + 2x_1(1)}{1 + 2(1)} = \frac{x_1^2+2x_1}{3} & 0 \leq x_1 \leq 1\\
1 & x_1 > 1
\end{cases}
$$
We set $x_2 = 1$ and the rule changed to $\frac{x_1^2+2x_1}{3}$. To hammer the point home, let's calculate $P(X_1 < 1/2| X_2 = x_2)$ for both the cases described above. For the case $x_2 = 0$

$$
P(X_1 < 1/2|x_2 = 0) = (1/2)^2 = 1/4
$$
and for the case $x_2 = 1$

$$
P(X_1 < 1/2|x_2 = 1) = \frac{(1/2)^2 + 2(1/2)}{3} = \frac{5/4}{3} = \frac{5}{12}
$$
Notice how the probabilities change given different values of $x_2$. Indeed, if we plot the cdfs for each a set of values of $x_2$, we shall also find that the cdfs look different as well. In the figure below, you can see the plot of $P(X_1 < x_1| X_2 = x_2)$ for values of $x_2$ equal to 0, 1/3, and 1, respectively. Note that we've ignored the plot above 1 and below 0, so you won't see it here. Each of the aforementioned conditional cdfs is represented by a different line pattern (the middle line is the cdf corresponding to $x_2 = 1/3$). Notice how they progress to a probability of 1 on a different path - dependent on the value of outcome of $X_2$.

```{r, echo = FALSE, fig.cap = "Changing cdfs!"}
cdf <- function(x1,x2) {
  p <- (x1^2 + 2*x1*x2)/(1 + 2*x2)
  p
}

x <- seq(0,1, by = 0.005)
probs <- c(cdf(x,0), cdf(x,1/3), cdf(x,1))
x2s <- rep(c('0','1/3','1'), each = 201)
plot_data <- data.frame(x, x2s, probs)

ggplot2::ggplot(plot_data, ggplot2::aes(x = x, y = probs, linetype = x2s)) +
  ggplot2::theme_bw() +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::geom_line() +
  ggplot2::labs(y = 'P(X1 < x1 | X2 = x2)', x = 'x1', linetype = 'Value of x2')

```
Thus, a good intuition for continuous random variables is similar to the one we built up for discrete random variables. The cdf of one continuous random variable changes with particular values of another random variable. 

:::

Now we direct out attention towards independence, and the convenience of working with independent random variables. Suppose we are given two **independent** random variables $X$ and $Y$ with cdfs $F_X(x)$ and $F_Y(y)$, respectively. I've used the subscripts on each cdf to make it extra clear to which random variable they belong. Further suppose we are given two events $A$ and $B$ in the space of $X$ and $Y$, respectively, and we would like to know $P(A \cap B)$. That is to say, we would like to know the probability of A *and* B occurring. Because our two random variables are independent - and therefore so are $A$ and $B$ - we just multiply their two respective probabilities together.

$$
P(A \cap B) = P(A)P(B)
(\#eq:indcont)
$$

where $P(A)$ and $P(B)$ would each be calculated using $F_X(x)$ and $F_Y(y)$, respectively. Does this look familiar? Indeed it is the same equation we used in the discrete case, and it is generally true regardless of the type of random variable. That is to say, if we are given two independent random variables - both discrete, one discrete, or both continuous - then we may use the equation above.

::: {.example}
Suppose we are working with two independent random variables - $X$ and $Y$ - each with the same pdf that looks like this

$$
f_X(a) = f_Y(a) = 
\begin{cases}
e^{-a} & a \geq 0 \\
0 & a < 0
\end{cases}
$$
Thus, for each function, we input a value of $a$ and it returns a density. In previous sections, we've seen that the cdf for $X$ and $Y$ is

$$
P(X < a) = F_X(a) = F_Y(x) =
\begin{cases}
1 - e^{-a} & a \geq 0\\
0 & a < 0
\end{cases}
$$
Now let's say that we are conducting a random experiment, and we are interested in the probability that $X < 1$ and $Y < 1$. Because these two random variables are independent, equation \@ref(eq:indcont) will work nicely.

$$
P({X < 1} \cap {Y < 1}) = P(X < 1)P(Y < 1) = (1-e^{-1})(1-e^{-1})
$$
We simply multiply the two probabilities together.
:::

### Expectation and Variance of a Continuous Random Variable.
To this point, we have seen that independence has a very similar interpretation for the continuous case. Indeed the similarities between discrete and continuous random variables do not end there! The expectation and variance are no different in this regard. Like the discrete case, the expectation of a continuous random variable can be thought of as the balancing point - or center of mass - of the distribution. If we put a wedge under the distribution at the value of its expectation, the pdf would balance perfectly. The variance, on the other hand, can also be viewed as a measure of dispersion. A larger variance means the probability density of f(x) is spread out more.

The calculation of a continuous expectation is a bit more complex, and requires integration techniques which are (again) beyond the scope of this book. However, the intuition for the calculation is very similar to the discrete case. We'll walk through it right now. Consider the probability density function

$$
f(x) = 
\begin{cases}
x & 0 \leq x \leq \sqrt{2} \\
0 & Everywhere\: else 
\end{cases}
$$

This probability density function is plotted below and it's not massively different from the first density we talked about. This time we're looking at a triangle region under the pdf (maybe take a second verify that $f(x)$ has all the required properties of a pdf).

```{r, echo = FALSE, fig.cap = "Triangle pdf"}
a <- data.frame(x = c(0,sqrt(2)), 
                y = c(0,0),
                x_end = c(sqrt(2),2.5),
                y_end = c(sqrt(2),0))

b <- data.frame(x = c(0,sqrt(2)), 
                y = c(0,sqrt(2)))

d <- data.frame(x = c(sqrt(2)), 
                y = c(0))
                

ggplot2::ggplot(a, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_segment(data = a, ggplot2::aes(x = x, y = y, xend = x_end, yend = y_end)) +
  ggplot2::scale_y_continuous(limits = c(0,2)) +
  ggplot2::scale_x_continuous(limits = c(0,2.5)) +
  ggplot2::labs(y = "Density", x = "x") +
  ggplot2::geom_point(data = b, ggplot2::aes(x=x,y=y)) +
  ggplot2::geom_point(data = d, ggplot2::aes(x=x,y=y), shape = 21, fill = 'white') +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::theme_bw()

```

The cdf of this probability density function is

$$
F(x) = 
\begin{cases}
0 & x < 0 \\
\frac{x^2}{2} & 0 \leq x \leq \sqrt{2} \\
1 & x > \sqrt{2} 
\end{cases}
$$

I've used integration techniques to get this equation, but you can derive it from the equation for the area of a triangle. Recall, that with regards to the discrete case, the expectation is calculated by multiplying each outcome of $X$ by its corresponding probability $p(x)$ and summing the results. That is, if there are $n$ outcomes in the space of $X$,

$$
E(X) = \sum_{i = 1}^{n}x_ip(x_i)
$$

Let's try something similar for our continuous random variable, first we'll divide the support of our random variable $X$ into four equal ranges - each with width $\frac{1}{4}\sqrt{2}$. Specifically,

$$
\{0 < X < \frac{1}{4}\sqrt{2}\}, \:
\{\frac{1}{4}\sqrt{2}\ < X < \frac{2}{4}\sqrt{2}\}, \:
\{\frac{2}{4}\sqrt{2}\ < X < \frac{3}{4}\sqrt{2}\}, \:
\{\frac{3}{4}\sqrt{2}\ < X < \frac{4}{4}\sqrt{2}\}
$$
where I haven't simplified the fractions to highlight the four equal regions. Next, we'll choose any value from each of the four ranges. Call each $x_i^*$ - where $i$ corresponds to range 1 through 4. For the sake of simplicity, let's select the middle value of each range - $\frac{1}{8}\sqrt{2} , \: \frac{3}{8}\sqrt{2}, \: \frac{5}{8}\sqrt{2}, \:$ and $\frac{7}{8}\sqrt{2}$, respectively. The ranges and our $x_i^*$s are depicted in the figure below.

```{r, echo = FALSE, fig.cap = "Triangle pdf divided into 4"}
f <- data.frame(x = c(sqrt(2) * 1/8, sqrt(2) * 3/8, sqrt(2) * 5/8, sqrt(2) * 7/8),
                y = c(0,0,0,0),
                labels = c('x*1','x*2','x*3','x*4'))

a <- data.frame(x = c(0,sqrt(2)), 
                y = c(0,0),
                x_end = c(sqrt(2),2.5),
                y_end = c(sqrt(2),0))

b <- data.frame(x = c(0,sqrt(2)), 
                y = c(0,sqrt(2)))

d <- data.frame(x = c(sqrt(2)), 
                y = c(0))

e <- data.frame(x = c(sqrt(2)/4, 2*sqrt(2)/4, 3*sqrt(2)/4, 4*sqrt(2)/4),
                y = c(0,0,0,0),
                x_end = c(sqrt(2)/4, 2*sqrt(2)/4, 3*sqrt(2)/4, 4*sqrt(2)/4),
                y_end = c(sqrt(2)/4, 2*sqrt(2)/4, 3*sqrt(2)/4, 4*sqrt(2)/4))
                

ggplot2::ggplot(a, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_segment(data = a, ggplot2::aes(x = x, y = y, xend = x_end, yend = y_end)) +
  ggplot2::geom_segment(data = e, ggplot2::aes(x = x, y = y, xend = x_end, yend = y_end)) +
  ggplot2::geom_point(data = f, ggplot2::aes(x = x, y=y)) +
  ggplot2::geom_text(data = f, ggplot2::aes(x=x, y=y, label= labels), nudge_y = .1) +
  ggplot2::scale_y_continuous(limits = c(0,2)) +
  ggplot2::scale_x_continuous(limits = c(0,2.5)) +
  ggplot2::labs(y = "Density", x = "x") +
  ggplot2::geom_point(data = b, ggplot2::aes(x=x,y=y)) +
  ggplot2::geom_point(data = d, ggplot2::aes(x=x,y=y), shape = 21, fill = 'white') +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::theme_bw()

```

As we have learned, the area under the pdf for a range corresponds to the probability that $X$ falls in that range. For example, the area under the pdf for the first range is $P(X < \frac{1}{4}\sqrt{2})$ and for the last range it is $P(\frac{3}{4}\sqrt{2} < X < \frac{4}{4}\sqrt{2})$. Using \@ref(eq:fundamental), we can calculate the latter as

$$
\begin{align}
P(\frac{3}{4}\sqrt{2} < X < \frac{4}{4}\sqrt{2}) &= P(X < \frac{4}{4}\sqrt{2}) - P(X < \frac{3}{4}\sqrt{2}) \\
&= F(\sqrt{2}) - F(\frac{3}{4}\sqrt{2}) \\
&= \frac{(\sqrt{2})^2}{2} - \frac{(3/4 \cdot \sqrt{2})^2}{2} \\
&= \frac{2}{2} - \frac{9/16 \cdot 2}{2} \\
&= 1 - \frac{9/8}{2} \\
&= 1 - \frac{9}{16} = \frac{7}{16}
\end{align}
$$

Putting it all together, we shall calculate the corresponding probability for each range, multiply each by it's corresponding $x_i^*$, and then sum the resulting values. In other words, call each range $R_i$ for simplicity, and

$$
\sum_{i = 1}^{4} x_1^*P(R_i)
$$
The equation above looks awfully similar to our equation for the expectation of a discrete random variable. Let's do the calculation for our four ranges and their corresponding $x_i^*$s.

$$
\begin{align}
\sum_{i = 1}^{4} x_1^*P(R_i) &= x_1^*P(R_1) + x_2^*P(R_2) + x^*_3P(R_3) + x^*_4P(R_4) \\
&= x_1^*\frac{1}{16} + x_2^*\frac{3}{16} + x_3^*\frac{5}{16} + x_4^*\frac{7}{16} \\
&= (\frac{1}{8}\sqrt{2})\frac{1}{16} + (\frac{3}{8}\sqrt{2})\frac{3}{16} +(\frac{5}{8}\sqrt{2})\frac{5}{16} + (\frac{7}{8}\sqrt{2})\frac{7}{16} \\
&=\sqrt2\frac{21}{32} = 0.92808
\end{align}
$$
I've skipped some of the computational details, but I imagine your overwhelming curiosity will lead you to calculate each step yourself. Using integral techniques, the exact expectation of our random variable can be calculated to be $\frac{2}{3}\sqrt{2}$, so our computation above is a pretty good approximation. What's neat is that we can make the approximation better by dividing the support of $X$ into smaller ranges and performing the calculation again with new $x_i$s selected from those ranges. In the table below, I've done just that. The column on the left indicates the number of ranges (1 to 1000), the second column from the left indicates the width of each of those ranges, and the third column indicates the resulting approximation of $E(X)$. Note that I've used the middle value of each range for $x_i$. The final column is the approximation's difference from the actual value (subtracting the actual from the approximation).

```{r, echo = FALSE, fig.cap = "Improving Approximations"}
apx <- function(n) {
  rng <- sqrt(2)/n
  mid <- rng/2
  x <- seq(0, sqrt(2), by = rng)
  probs <- (x^2)/2
  probs <- diff(probs)
  xi <- (x + mid)[1:length(x)-1]
  E <- t(xi) %*% (probs)
  vals <- list(n, rng, E)
  }

x <- c(1:10,20,50,100,1000)
diff <- c(apx(1)[[2]],apx(2)[[2]],apx(3)[[2]], apx(4)[[2]], apx(5)[[2]], apx(6)[[2]], 
          apx(7)[[2]],apx(8)[[2]],apx(9)[[2]],apx(10)[[2]], apx(20)[[2]], apx(50)[[2]],
          apx(100)[[2]], apx(1000)[[2]])
E <- c(apx(1)[[3]],apx(2)[[3]],apx(3)[[3]], apx(4)[[3]], apx(5)[[3]], apx(6)[[3]], 
          apx(7)[[3]],apx(8)[[3]],apx(9)[[3]],apx(10)[[3]], apx(20)[[3]], apx(50)[[3]],
          apx(100)[[3]], apx(1000)[[3]])

act <- rep(sqrt(2)*2/3, times = 14)
table <- data.frame(round(x,5),round(diff,5), round(E,5), round(E - act,5))
names(table) <- c('Number of Ranges', 'Range Width', 'Approximate E(X)', 'Difference from Actual')
x <- flextable::flextable(table)
x <- flextable::autofit(x)
x <- flextable::align(x, align = "center", part = 'body')
x <- flextable::hline(x)
x <- flextable::bg(x, bg = 'gray', part = 'header')
x

```

With more ranges - and smaller range widths - the value of our approximation gets closer and closer to the actual value. This observation leads us to an intuitive interpretation of how $E(X)$ is calculated for a continuous random variable. Suppose we were to make the ranges extremely small (and the number of ranges very high) and perform the computation above. That is, we take the probability that $X$ falls in each of our very small ranges, multiply each by a value of $x$ in that extremely small range, and then sum the resulting values. With a set of infinitesimally small ranges, this would give us $E(X)$. In practice, this operation of taking smaller and smaller ranges, multiplying, and summing is done via the methods of integration.

Indeed, you'll note that the procedure we have just walked through is very similar to the calculation for discrete random variables. Now we'll take it one step further. Recall how we could calculate the expectation of some function of a discrete random variable \@ref(eq:discreteexpectationfunction). For example, suppose were were working with a discrete random variable, and we are interested in the expectation of $f(X) = X + 1$, a.k.a. $E(f(X))$. For the sake of simplicity, let's say our discrete random variable has the pmf

$$
P(X = x) = p(x) = 
\begin{cases}
1/2 & x = 0 \\
1/2 & x = 1  
\end{cases}
$$
Then we would apply the function to each outcome in the space of $X$ and calculate the expectation with the resulting values. That is

$$
f(0)p(0) + f(1)p(1) = (0 + 1)1/2 + (1 + 1)1/2 = 3/2
$$
Now let's work towards an intuition of the computation of $E(f(x))$ in the continuous case. You will also see (intuitively speaking) how similar this computation is to the discrete case. Suppose our variable $X$ were continuous, and we were interested in $E(f(X))$. To estimate this expectation, we would first create $n$ ranges over the space of $X$. Here we'll refer to the ranges with $R_i$ - with $i$ indicating the ith range. Then we take an $x^*_i$ from each range, and apply the function to each - $f(x^*_i)$. As in the discrete case (and our discussion before that), the final step would be multiplying each $f(x^*_i)$ by the probability that $X$ falls in the ith range - $P(R_i)$ - and sum the resulting values. Put more succinctly:

$$
\sum_{i = 1}^{n} f(x^*_i)P(R_i)
$$

where we calculate $P(R_i)$ using the cdf of $X$. If we perform this process with very very (extremely) small ranges, then we will find the value of $E(f(X))$. Indeed, in practice, the aforementioned operation is achieved via integration. Note the similarity to the discrete case \@ref(eq:discreteexpectationfunction). Ultimately the interpretation is also the same: $E(f(X))$ is the center of mass of $f(X)$ after the function has been applied to each outcome in the space of $X$.

Given the similarities between the continuous and discrete case, it may therefore be unsurprising (although no biggie if it is surprising) to learn that the expectation of a continuous random variable has the same previously discussed properties as that of a discrete random variable. That is 

::: {.theorem #expcont name="Some Properties of the Continuous Expectation"}
If $a$ and $b$ are constants, then

1. $E(b) = b$

2. $E(aX) = aE(X)$

3. More generally, $E(aX + b) = aE(X) + E(b)$

If $X_1$ and $X_2$ are random variables, then 

4. $E(aX_1 + bX_2)$ = $aE(X_1) + bE(X_2)$
:::

And, continuous random variables also share the previously discussed properties of the discrete variance. That is,

::: {.theorem #contvar name="Some Properties of the Continous Variance"}
If $a$ and $b$ are constants, then

1. $Var(b) = 0$

2. $Var(aX) = a^2Var(X)$

3. More generally, $Var(aX + b) = a^2Var(X) + Var(b) = a^2Var(X) + 0$

If $X_1$ and $X_2$ are random variables, **and they are independent**, then 

4. $Var(aX_1 + bX_2)$ = $a^2Var(X_1) + b^2Var(X_2)$
:::

In light of the above equivalence, we will heretofore simply refer to the two theorems above as properties of the expectation and variance, respectively, rather than belonging to some class of discrete or continuous random variables. In other words, we may use the properties of the expectation and variance for random variables generally (if those random variables have expectations and variance; we'll look at an example of a random variable that doesn't have these in the next section).

::: {.example}
Let's work with the random variable $X$ that we defined this section. As stated above, it has $E(X) = \frac{2}{3}\sqrt{2}$. Suppose we are given a random variable - $Y \sim B(10,1/2)$ - and we would like to know $E(X + Y)$. Our properties make this calculation a breeze.

$$
E(X + Y) = E(X) + E(Y) = \frac{2}{3}\sqrt{2} + 10\frac{1}{2} = \frac{2}{3}\sqrt{2} + 5
$$
:::

### Practice Problems

Consider the pdf and cdf below for the random variable $X$ and do your best to calculate the probabilities in questions 1 through 6.

$$
f(x) = \begin{cases}
0 & x < -1/2 \\
1  & -1/2 \leq x \leq 1/2 \\
0 & x > 1/2
\end{cases}
$$

$$
P(X < x) = F(x) = \begin{cases}
0 & x < -1/2 \\
x + 1/2 & -1/2 \leq x \leq 1/2 \\
1 & x > 1/2
\end{cases}
$$

1. $P(X < 0)$

2. $P(X < 1/2)$

3. $P(-1/4 < X < 1/8)$

4. $P(\{X > 1/4\} \: \cup \: \{X < -1/4\})$

5. $P(\{X > 1/4\} \: \cap \: \{X < -1/4\})$

6. $P(X < 10000)$

Consider the pdf and cdf below for the random variable $X$ and do your best to calculate the probabilities in questions 7 through 10. Note: Don't feel the obligation to calculate $e^-x$. Just leave it in that form.

$$
f(x) = \begin{cases}
0 & x < 0 \\
e^{-x} & x \geq 0
\end{cases}
$$

$$
P(X < x) = F(x) = \begin{cases}
0 & x < 0 \\
1-e^{-x} & x \geq 0
\end{cases}
$$

7. $P(0 < X < 1 \: \cup \: X > 2)$

8. $P(X > 1)$

9. $P(1 < X < 4 \: \cap \: X > 3)$

10. $P(X = 1)$ (Hint: $X = 1$ may also be written using $\leq$ inequalities)

11. Suppose $P(X > x) = e^{-x/2}$. What is the cdf of $X$?

12. Suppose we had three random variables - $X$, $Y$, and $Z$ - each with the pdf/cdf of questions 1 through 6. On a given random experiment, what is the probability that $X > 1/4$, $Y < 1/4$, and $Z < 1/4$? What is the probability that exactly two of the three outcomes are above 1/4?

13. Consider the probability density for the random variable $X$ below. Draw the plot of the density. Based on the plot what is $E(X)$? Why?


$$
f(x) = \begin{cases}
0 & x < 0 \\
1 & 0 \leq x \leq 1 \\
0 & x > 1
\end{cases}
$$

14. Consider the probability density for the random variable $Y$ below. Draw the plot of the density. Based on the plot what is $E(Y)$? Does this variable have a larger or smaller variance than the variable in question 12? Why or why not?

$$
f(y) = \begin{cases}
0 & y < 0 \\
1/2 & 0 \leq y \leq 2 \\
0 & y > 2
\end{cases}
$$

Suppose we are given two continuous random variables - $X_1$ and $X_2$. Each with $E(X_1) = 4$ and $E(X_2) = 2$, and variances of $1$ and $3$, respectively. Using the properties you've learned in this section, try your best to answer the questions below. Note if the properties learned here do not apply.

15. What is $E(X_1) + E(X_2)$?

16. What is $Var(2X_2)$?

17. What is $Var(X_1 + X_2)$?

18. What is $E(X_1) + E(Y)$ where $Y$ is a poisson random variable with $\lambda = 5$.

### Practice Problem Answers

1. 1/2
2. 1
3. 3/8
4. 1/2
5. 0
6. 1
7. $e^{-1} + e^{-2}$
8. $e^{-1}$
9. $e^{-3}-e^{-4}$
10. 0 (WEIRD!)
11. $1 - e^{-x/2}$
12. 9/64, 9/64
13. 1/2; Why? Because if you put a wedge right under 1/2, it looks like the plot would balance.
14. 1; Yes, it does have a larger variance than the variable in question 13 because its probability density is much more spread out.
15. 6
16. 4
17. The properties we learned in this section won't work for this one. The two variables do not appear to be independent.
18. 9

## A Very Special Distribution (and Some Other Cool Distributions Too)
We have spent the last couple of sections talking generally about continuous random variables and their distributions. For the final topic of this chapter, we are going to turn our attention towards some commonly occurring probability distributions. The first (and arguably most prolific) distribution we will discuss is the **normal distribution**. As you will see, we will - for various reasons - work frequently with normal distributions in our statistical adventures. Our discussion of the normal distribution shall proceed in two steps. First we will talk about the shape of the normal distribution and some of it's basic properties. Then we will talk about calculating probabilities for a normally distributed random variable. After that, we'll extend our discussion to other important distributions, although, admittedly, with a little less depth.

### The Normal Distribution

The normal density - or normal distribution - is often referred to as a "bell curve" because of its shape. I've plotted a normal distribution in the figure below. Notice how it has a maximum (a peak) at it's center, and the density declines the farther values are from this middle point. Now, supposing we put a wedge right under the value where that peak occurs. The distribution would balance perfectly. That is to say, the value of $x$ at the center of the normal distribution (where $f(x)$ is at a maximum) is the center of mass or expectation of the normally distributed random variable. Also note how the normal distribution is *symmetric* about its center of mass. At a more intuitive level, this means that the function to the left of the center is the mirror image of the function to the right. More specifically: given some value - $a$ - then $f(\mu - a) = f(\mu + a)$ - where $f(x)$ is the normal pdf (recall $\mu$ is just another way of saying $E(X)$). In other words, the value of $f(x)$ - our probability density - is the same at equivalent deviations above and below its center of mass. 

Next, scan the normal distribution plot from left to right. Notice the curve on the left hand side all the way to the plotted point at $(-1,f(-1))$ opens up like a $\cup$. In other words, the curve is *convex* to this point. However, after $(-1,f(-1))$ the curve opens down towards the $x$-axis. That is to say, it changes to *concave*. When a curve switches from convex to concave (or vice versa), we call it an **inflection point**. In our example figure below, the normal distribution has two such inflection points at $(-1,f(-1))$ and $(1,f(1))$, respectively. In the second, the curve switches from concave to convex and remains that way thereafter.

```{r, echo = FALSE, fig.cap = "The Normal Distribution"}
x <- seq(-2.5, 2.5, by = .01)
y <- dnorm(x)
z <- data.frame(x,y)

ggplot2::ggplot(z, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::geom_vline(xintercept = 0) +
  ggplot2::geom_point(x = -1, y = dnorm(-1)) +
  ggplot2::geom_point(x = 1, y = dnorm(1)) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = 'x', y = 'Probability Density')

```
In fact, all normal distributions have exactly two inflection points. Suppose $\sigma$ is the standard deviation of our normally distributed random variable (recall the standard deviation is the square root of the variance). The inflection points of a normally distributed random variable are located at exactly - $x = \mu \pm \sigma$. In words, they are located at one standard deviation above or below the mean/expectation of the normally distributed random variable.

Just by looking at the figure above, we have already learned a great deal about the normal distribution. Namely:

1. The normal distribution has its expectation at its peak/center;
2. it is symmetric;
3. and it has inflection points at $x = \mu \pm \sigma$

Keep these properties in mind as we delve deeper into the mathematics of the normal distribution. The the equation for the probability density function of a normally distributed random variable is a bit complex, but knowing it will be useful for our discussion. If $X$ is a normally distributed random variable, then it's pdf is

$$
f(x) = \frac{1}{\sqrt{2\pi}\sigma}e^{-\frac{1}{2}\left[\frac{x - \mu}{\sigma}\right]^2}, \:\:-\infty < x < \infty
$$
and the space of $X$ is negative infinity to positive infinity (all the real numbers). In more mathy (technical term) textbooks, the pdf is often written

$$
f(x) = \frac{1}{\sqrt{2\pi}\sigma} exp \left [-\frac{1}{2}\left(\frac{x - \mu}{\sigma}\right)^2\right], \:\:-\infty < x < \infty
(\#eq:normaldensity)
$$

where the factor to the right of and including "*exp*" is just another way of writing $e^{-\frac{1}{2}\left[\frac{x - \mu}{\sigma}\right]^2}$. Notice the two values $\mu$ and $\sigma$. These two values are the parameters of the normal distribution. The value of $\mu$ is the middle point of the distribution (below the peak) and the value of $\sigma$ determines its spread. In the figure below, you can see several examples of normal distributions with various selections of $\mu$ and $\sigma$

```{r, echo = FALSE, fig.cap = "Different locations and spread"}
x1 <- seq(-4,4, by = .01)
x2 <- seq(-8,8, by = .01)
x3 <- seq(-9,9, by = .01)
x4 <- seq(1,7, by = .01)
x5 <- seq(-1,9, by = .01)
x6 <- seq(-4,12, by = .01)
y1 <- dnorm(x1)
y2 <- dnorm(x2, sd = 2)
y3 <- dnorm(x3, sd = 3)
y4 <- dnorm(x4, mean = 4, sd = 1)
y5 <- dnorm(x5, mean = 4, sd = 2)
y6 <- dnorm(x6, mean = 4, sd = 3)
x <- c(x1,x2,x3,x4,x5,x6)
y <- c(y1,y2,y3,y4,y5,y6)
mu <- rep(c("mu = 0","mu = 0","mu = 0",
              "mu = 4","mu = 4","mu = 4"), 
            c(length(x1), length(x2), length(x3),
              length(x4), length(x5), length(x6)))
sigma <- rep(c('sigma = 1', 'sigma = 2', 'sigma = 3',
            'sigma = 1', 'sigma = 2', 'sigma = 3'), 
          c(length(x1), length(x2), length(x3),
            length(x4), length(x5) , length(x6)))
z <- data.frame(x,y, mu = as.factor(mu), sigma = as.factor(sigma))

ggplot2::ggplot(z, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::scale_x_continuous(breaks = c(-8,-6,-4,-2,0,2,4,6,8,10)) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = 'x', y = 'Probability Density') +
  ggplot2::facet_grid(cols = ggplot2::vars(sigma), rows = ggplot2::vars(mu))

```
Note how the normal distributions in the top row of the figure all have $\mu = 0$, but their values of $\sigma$ differ. From left to right, the distributions have $\sigma$ equal to 1, 2, and 3, respectively. Without any special math, you can see how the distribution in the first column is less spread out than the distributions in the second and third columns. Now take a look at the second row. Each distribution in this row has the same value of $\sigma$ as the distribution in the same column above. However, note how they are moved slightly to the right on the number line. THis is because these distributions each have $\mu = 4$ as opposed to the $\mu = 0$ in the first row. Their peaks/centers are therefore located at 4 as opposed to 0.

Thus, the parameters of $\mu$ and $\sigma$ control the *location* and the *scale* of the normal distribution. A normal distribution with $\mu = 10$ has a center/peak 15 points to the right of a distribution with $\mu = -5$, and a distribution with $\sigma = 5$ is more spread out than one with $\sigma = 1$. You may have noticed, based on our discussion here and our discussions in previous sections, that the parameters $\mu$ and $\sigma$ are very similar to the expectation and variance. One determines the center of the normal distribution and the other determines its dispersion, spread, or scale.

Indeed, your intuition in this situation would be correct (although if you didn't get such an intuition, that's cool too). If $X$ is a normally distributed random variable, then

$$
E(X) = \mu \\
Var(X) = \sigma^2
$$

Thus $\mu$ is the expected value of $X$ and $\sigma$ is the standard deviation. Indeed it is not a coincidence that the two parameters of the normal distribution are (by convention) referred to as $\mu$ and $\sigma$ - which are symbols used more generally to refer to the expectation and standard deviation of *any* random variable. Sometimes when referring to a normally distributed random variable we will use the symbol $X \sim N(\mu, \sigma^2)$ as a shorthand. Important note: the variance - $\sigma^2$ - is used in this notation, not the standard deviation.

Let's now take a more in depth look at the normal pdf. Direct your attention to the exponent of $e$ in the equation. 

$$
exp \left [-\frac{1}{2}\left(\frac{x - \mu}{\sigma}\right)^2\right]
$$
We shall learn a lot about the normal distribution from the term $\frac{x - \mu}{\sigma}$. This term says that to determine the density at some outcome $x$ we first subtract the mean/expectation of $X$ from that value and divide by it's standard deviation. That is to say, we calculate the number of standard deviations the particular value - $x$ - is from the mean of the normally distributed random variable $X$. For a given value of $x$, it is often convention to call this value a **z-score**.

::: {.example}
Suppose we have a normally distributed random variable $X \sim N(10,4)$. We would like to know the z-score of the value 12. First, we are given a *variance* of 4, so we will have to take the square root of the variance to get the standard deviation. Next

$$
Z = \frac{x - \mu}{\sigma} = \frac{12 - 10}{2} = 1
$$
:::

Now, suppose we do the $Z$ score calculation on every value in the space of a normally distributed random variable $X$. That is to say, we create a new random variable that has z-scores as its outcomes. 

$$
Z = \frac{X - \mu}{\sigma}
$$
You may be thinking: What the heck!? We have never discussed creating a *new* random variable before. But we have. We just never called it that. Recall, for example, our discussions of the expectation of discrete random variables. At one point we talked about calculating the expectation of some $f(X)$. In particular, we discussed the following discrete distribution

$$
p(x) = 
\begin{cases}
1/5 & x_1 = 1  \\
1/5 & x_2 = 2  \\
1/5 & x_3 = 3  \\
1/5 & x_4 = 4  \\
1/5 & x_5 = 5 
\end{cases}
$$
And we sought to calculate the expectation of $E(f(X))$, where $f(X) = 2X$. In words, we wanted to know the expectation of a function that multiplies every value in the space of the random variable $X$ by 2. Thus, we applied this function to every value in the space of $X$ and

$$
p(x) = 
\begin{cases}
1/5 & x_1 = 1 & f(1) = 2(1) = 2 \\
1/5 & x_2 = 2 & f(2) = 2(2) = 4 \\
1/5 & x_3 = 3 & f(3) = 2(3) = 6 \\
1/5 & x_4 = 4 & f(4) = 2(4) = 8 \\
1/5 & x_5 = 5 & f(5) = 2(5) = 10 
\end{cases}
$$

I did not mention it at the time, but what we actually did was create a *new* random variable - one with the space of ${2,4,6,8,10}$ - each assigned a probability of 1/5. That is to say, we created a new random variable - call it $Y$ - and

$$
Y = 2X
$$

Albeit, the process of creating a new continuous random variable is a bit more mathematically complicated, but the intuition is the same.

Now returning to our normally distributed random variable - $X \sim N(\mu, \sigma^2)$. We shall create a *new* distribution of z-scores. Put another way, if we were to conduct a random experiment with this new random variable, we would get a $z$ in standard deviations, rather than the particular value of $x$. The first thing I will tell you (and you shall have to accept) is that this new variable will also be normally distributed. However, to fully specify this normal distribution, we shall need values of $\mu$ and $\sigma^2$. That is we will need to determine the expectation and variance of $Z$. Recall that $Z$ is a function of $X$, and we have already, on several occasions learned how we might calculate the expectation of a function of $X$.

Specifically, we will use the properties of the expectation and variance we learned in the previous chapters/sections. With regards to $\mu = E(Z)$,

$$
\begin{align}
E(Z) &= E(\frac{X - \mu}{\sigma}) \\
&= \frac{1}{\sigma}E(X - \mu) \\
&= \frac{1}{\sigma}\left[E(X) - E(\mu)\right] \\
\end{align}
$$

Because $\sigma$ is a constant. Now $\mu$ is also not a random variable. It's a constant, so $E(\mu) = \mu$, and furthermore, $E(X) = \mu$. Thus,

$$
\frac{1}{\sigma}\left[\mu - \mu\right] = 0
$$

The mean of our new z-score random variable is therefore zero. But what about the variance? Well, recalling the properties of the variance

$$
\begin{align}
Var(Z) &= Var(\frac{X - \mu}{\sigma}) \\
&= \frac{1^2}{\sigma^2}Var(X - \mu) \\
\end{align}
$$

Again, notice that $\mu$ is a constant, so using the properties of the variance $Var(X - \mu) = Var(X)$. And also note that the variance of our normally distributed random variable $X$ is $\sigma^2$.

$$
\begin{align}
&= \frac{1}{\sigma^2}Var(X - \mu) \\
&= \frac{1}{\sigma^2}Var(X) \\
&= \frac{1^2}{\sigma^2} \sigma^2 = 1\\
\end{align}
$$

Thus, $Z$ - our distribution of z-scores - is a normally distributed random variable with $\mu = 0$ and $\sigma^2 = 1$. If we enter these values into the equation for the normal distribution, we get

$$
f(x) = \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}\left(z\right)^2}, \:\:-\infty < x < \infty
(\#eq:standardnormal)
$$

which we will call the **standard normal distribution**, a.k.a. $Z \sim N(0,1)$. This is the distribution we get when we calculate a z-score for all values in the space of a normally distributed random variable. Now, I realize that the term "standard normal" sounds a bit redundant, and while I don't always espouse convention, you'll just have to put up with it here.

In practice, we will frequently work with the standard normal distribution. Indeed, when we are hoping to calculate probabilities for a normally distributed random variable, we will use its cdf. 

### The Normal Cumulative Distribution Function
Although the normal distribution (or normal density) is interesting in of itself, we will frequently work with its cumulative distribution function. Recall in our previous work with cumulative distribution functions, we used an equation to calculate probabilities. We would enter some value into this equation for the cdf, do some computation, and get $P(X < x)$ - where the lower case $x$ is the value we entered into the equation. Sadly, no such function for a normal distribution exists. This (thankfully) does not mean we can't calculate the area under the normal distribution (a.k.a., integrate). It just means that we have to calculate the area by other means. 

In the absence of an equation, there are many modern and not so modern ways to describe the normal cumulative distribution function. For our purposes we will use a table of values (more on that soon). You may also choose to use your favorite statistical computing program (like R or Python). The aforementioned table helps us take a given *z*-score and lookup $P(Z < z)$. Put another way, it describes the cumulative distribution of the standard normal distribution - the distribution of *z*-scores we calculated just before this discussion. Although we do not have a standard equation, the cdf function is often referred to as

$$
\Phi(z) = P(Z < z) 
(\#eq:normalcdf)
$$

which is the function that our table of values describes numerically. 

```{r, echo = FALSE, fig.cap = "The Standard Normal PDF and CDF"}

x_1 <- seq(-3,3,by = 0.01)
y_1 <- dnorm(x_1)
x_2 <- seq(-3,3,by = 0.01)
y_2 <- pnorm(x_1)
pdf <- data.frame(x = x_1, y = y_1, grp = 'The Standard Normal pdf')
cdf <- data.frame(x = x_2, y = y_2, grp = 'The Standard Normal cdf')
all_data <- rbind(pdf,cdf)
all_data$grp <- factor(all_data$grp, levels = c(
  'The Standard Normal pdf', 'The Standard Normal cdf'
), ordered = TRUE)

ggplot2::ggplot(all_data, ggplot2::aes(x = x, y = y, color)) +
  ggplot2::geom_line() +
  ggplot2::facet_grid(cols = ggplot2::vars(grp)) +
  ggplot2::labs(x = 'z', y = '') +
  ggplot2::theme_bw() 

```

We will first walk through using the standard normal cdf table, and then we'll do a few examples. Suppose we want to determine $P(Z < 0.32)$. The table below presents the first few rows of the *z*-table that we will use for our work. The first column of the table represents the first digit and first decimal of our *z*-value. In our case, we would find the row in this column with the value of 0.3. Each column after the first represents a value of the second decimal place (a.k.a., the hundredths place) from 0.00 all the way to 0.09. Our *z*-value has a 2 in it's second decimal place, so we would scan to the right and use the value under the 0.02 column header. In this case, the value in the cell 0.62552. That is to say, $P(Z < 0.32) = 0.62552$. I've also highlighted the value (with a nice gentle teal) in the table to make it a bit easier to see what we are talking about.

```{r, echo = FALSE, cache = TRUE}


table <- data.frame(z = seq(0,.5,by = .1),
           '0.00' = round(pnorm(seq(0,.5,by = .1)),5),
           '0.01' = round(pnorm(seq(0.01, 0.51,by = .1)),5),
           '0.02' = round(pnorm(seq(0.02, 0.52,by = .1)),5),
           '0.03' = round(pnorm(seq(0.03, 0.53,by = .1)),5),
           '0.04' = round(pnorm(seq(0.04, 0.54,by = .1)),5),
           '0.05' = round(pnorm(seq(0.05, 0.55,by = .1)),5),
           '0.06' = round(pnorm(seq(0.06, 0.56,by = .1)),5),
           '0.07' = round(pnorm(seq(0.07, 0.57,by = .1)),5),
           '0.08' = round(pnorm(seq(0.08, 0.58,by = .1)),5),
           '0.09' = round(pnorm(seq(0.09, 0.59,by = .1)),5)
           )

names(table) <- c('z', '0.00','0.01','0.02','0.03','0.04','0.05','0.06',
                  '0.07','0.08','0.09')

x <- flextable::flextable(table)
x <- flextable::autofit(x)
x <- flextable::align(x, align = "center", part = 'all')
x <- flextable::hline(x)
x <- flextable::vline(x, j = 1)
x <- flextable::bg(x, bg = 'gray', part = 'header')
x <- flextable::bg(x, bg = 'gray', j = 1)
x <- flextable::bg(x, i = 4, j = 4, bg = '#B2EBF2')
x <- flextable::bold(x, i = 4, j = 4)
x

```

::: {.example}
Suppose we are asked to determine $P(X < 3)$ for $X \sim N(2,4)$. Based on that fancy notation, we know that $X$ is a normally distributed random variable - a perfect situation for the standard normal cdf. We will first have to convert our particular value of $x = 3$ into a *z*-score, and then we can figure out the value of $\Phi(z)$.

First note that we are given $\mu = 2$ and $\sigma^2 = 4$. Thus, the standard deviation of $X$ is $\sqrt{4} = 2$. Using this information, let's calculate the *z*-score.

$$
\frac{3 - \mu}{\sigma} = \frac{3-2}{2} = \frac{1}{2}
$$
Now we need to use our fancy table to determine the value of

$$
\Phi\left(\frac{3 - \mu}{\sigma}\right) = \Phi\left(\frac{1}{2}\right)
$$
Our *z*-score has a value of 0.5, so we will go down to the row in the table corresponding to this value. The hundredths place is zero, so we will take the very first value in the row - 0.69146. That is $P(X < 3) = 0.69146$. The distribution of $X$ - with the aforementioned area $P(X < 3)$ highlighted (in a very serene teal) - can be found in the figure below.

```{r, echo = FALSE}
x <- seq(-4,8, by = 0.01)
y <- dnorm(x,2,2)
func_vals <- data.frame(x,y)
x_0 <- seq(-4,3, by = 0.01)
y_0 <- rep(0, times = length(x_0))
y_1 <- dnorm(x_0,2,2)
lin_vals <- data.frame(x_0,y_0,y_1)

ggplot2::ggplot(func_vals, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_segment(data = lin_vals, 
                        ggplot2::aes(x = x_0, y = y_0, xend = x_0, yend = y_1),
                        color = '#B2EBF2') +
  ggplot2::geom_line() +
  ggplot2::geom_segment(x = 3, y = 0, xend = 3, yend = dnorm(3,2,2)) +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::labs(y = 'Probability Density', x = 'x') +
  ggplot2::geom_text(x = 1.1, y = 0.075, label = 'P(X < 3)', size = 5) +
  ggplot2::theme_bw()
```
:::

In the example above, we worked with a value that was greater than its mean. Supposing 